Separable Wavelet Scattering

### 2024-05-06 dissertation/content/paper2/main.tex
consolodated references

\section{Introduction}
Underwater \ac{pam} systems are an effective non-invasive tool to monitor marine activity, with many research endeavours to automate cetacean monitoring and detection over the last decades. \Ac{ml} is a particularly prevalent approach to detecting and classifying cetacean calls from \ac{pam} audio \citep{MLreview}, although some recent publications remain focused on traditional approaches that seem to oppose \ac{ml} trends \citep{jacques1, jacques2}. The disparity between the approaches are often due to a lack of data or a high uncertainty of the data quality \citep{casey2019}, which are detrimental to \ac{nn} and deep learning models.

@@

Large annotated datasets may not necessarily be accessible to automate detection and classification in conservation studies, which motivates the further improvement of ``white-box" strategies as opposed to using \acp{nn}. For \ac{poi} detection purposes, \ac{bled}, the \ac{se} detector and the spectrogram correlator \citep{casey2017} are mostly used in practise, often with software such as PAMGuard \citep{PAMGuard}. These detectors require the user to choose many different hyperparameters for the purpose of adaptivity to noise conditions and signal level. They do not necessarily classify signals, and instead pinpoint \acp{poi}, which suffer from many false positive detections. On the other hand, very accurate \ac{nn} approaches exist for a variety of species \citep{alexnet_killer_whales, cnn_multiple_whale_classes, narw_cn_denoising, cnn_fin_whale}, but require large datasets and careful data preparation.

Non-\ac{nn} approaches have also obtained very accurate results, most of which utilise \acp{mfcc} as features \citep{MFCC_HMM_birds, mfcc_hmm_brydes}. A \ac{hmm} is deployed in the aforementioned studies, thereby allowing the modelling of time-dependent sound characteristics without requiring a \ac{nn}.

@@

Automated classification is often the end goal of many \ac{pam} systems, but identifying \acp{poi} without classification is a necessary step to assist in the annotation and sound exploration process, especially when surveying new and unknown oceanic regions. As such, generalised detection is often performed with \acp{bled} and \ac{se} detectors \citep{se_dolphin}. \Ac{se} detectors have consistently shown better performance than energy detectors \citep{se_erbe_king}. The previous approach by \citet{mypaper} modifies \ac{se} detectors to stabilize the entropy measure and provide pseudo-probabilistic output, thereby improving endpoint detection accuracy and the interpretability of the detector's output. \ac{se} detectors naturally motivate extension by including feature extraction and classification using the same \ac{tf}-decomposition used for \ac{se} calculation, ultimately reducing computational resources.

@@

Figure \ref{fig:det_roc} shows the \ac{roc} curve for detection on a sample-by-sample basis. Incremental performance improvement is demonstrated by adding a \ac{gmm} and enhancing the detector with whitening. In Figure \ref{fig:det_calls}, we show that the proposed detector is able to detect the majority calls while detecting fewer noise.  As illustrated in many previous studies, \ac{se} greatly outperforms a traditional \ac{bled} detector \citep{se_erbe_king}. Whitening is also shown to greatly improve the capabilities of the \ac{bled} detector.

@@

As seen in Figures \ref{fig:det_roc} and \ref{fig:det_calls}, the proposed \ac{se} detector outperforms the previous method \citep{mypaper}. Typical \ac{bled} with a fixed threshold performs very poorly due to varying noise conditions. \ac{bled} has repeatedly been shown to not be a reliable \ac{poi} detector \citep{jacques1, se_erbe_king}.

@@

% \begin{table}[h!]
% \caption{Number of labeled detections for all vocalisations of interest. The detections are produced by the whitened \ac{gmm} \ac{se} detector. A significant number of detections are discarded in the training data due to overlapping.}
% \label{tab:numdetections}
% \begin{ruledtabular}
% \renewcommand{\arraystretch}{0.8}
% \begin{tabular}{lc} 
% \textbf{Vocalisation}      & \textbf{Number of detections}  \\ \hline
% Bm\_Ant\_A & 1514              \\
% Bm\_Ant\_B & 474              \\
% Bm\_Ant\_Z & 91               \\
% Bm\_D      & 328               \\
% Bp & 50                \\
% Noise      & 25638          \\
% Overlapping & 150 \footnotemark[1]      \\ 
% \end{tabular}
% \end{ruledtabular}
% \footnotetext[1]{Overlapping detections contain at least 2 overlapping annotations of different classes. This implies that more than 300 vocalisations are discarded as a result of overlap, thereby drastically reducing training data for the \textit{Bm\_D} and \textit{Bp\_20Plus} classes, which have a higher tendency for overlap.}
% \end{table}

@@

Our disposal of the time dimension by using the maximum naturally lends our classification model being replaced with a \ac{hmm} to incorporate time-dependent relationships similar to that of \citet{mfcc_hmm_brydes} and \citet{dmd_hmm}. However, instead of classifying on a frame-by-frame basis such as in \citet{otherScattering, ws_audio, ws_ecg}, we utilise the scattering coefficients to first detect a variable-length segment of audio which may contain whale vocalisations, thereby reducing the amount of data in the noise class. This provides a distinct advantage over other methods, in that it already incorporates the extracted feature coefficients in detection prior to classification. As such, large portions of noise are discarded, the amount of which is tuned by the probability threshold of the proposed \ac{se} detector. This contrasts to \citet{casey2019}, which contains over 3 million noise class samples for 5000 vocalisation samples using the sliding window approach. This corresponds to a ratio of 1:600 for vocalisation to noise classes, whereas our method, which includes a detector, reduces this ratio to approximately 1:7 (all classes), at the cost of rejecting very low \ac{snr} vocalisations.

>>>>

\section{Introduction}
Underwater \ac{pam} systems are an effective non-invasive tool to monitor marine activity, with many research endeavours to automate cetacean monitoring and detection over the last decades. \Ac{ml} is a particularly prevalent approach to detecting and classifying cetacean calls from \ac{pam} audio \citep{MLreview}, although some recent publications remain focused on traditional approaches that seem to oppose \ac{ml} trends \citep{internalsignaldetection, jacques2}. The disparity between the approaches are often due to a lack of data or a high uncertainty of the data quality \citep{casey2019}, which are detrimental to \ac{nn} and deep learning models.

@@

Large annotated datasets may not necessarily be accessible to automate detection and classification in conservation studies, which motivates the further improvement of ``white-box" strategies as opposed to using \acp{nn}. For \ac{poi} detection purposes, \ac{bled}, the \ac{se} detector and the spectrogram correlator \citep{casey2017} are mostly used in practise, often with software such as PAMGuard \citep{PAMGuard}. These detectors require the user to choose many different hyperparameters for the purpose of adaptivity to noise conditions and signal level. They do not necessarily classify signals, and instead pinpoint \acp{poi}, which suffer from many false positive detections. On the other hand, very accurate \ac{nn} approaches exist for a variety of species \citep{MLalexnet, cnn_multiple_whale_classes, narw_cn_denoising, cnn_fin_whale}, but require large datasets and careful data preparation.

Non-\ac{nn} approaches have also obtained very accurate results, most of which utilise \acp{mfcc} as features \citep{MFCC_HMM_birds, hmm2}. A \ac{hmm} is deployed in the aforementioned studies, thereby allowing the modelling of time-dependent sound characteristics without requiring a \ac{nn}.

@@

Automated classification is often the end goal of many \ac{pam} systems, but identifying \acp{poi} without classification is a necessary step to assist in the annotation and sound exploration process, especially when surveying new and unknown oceanic regions. As such, generalised detection is often performed with \acp{bled} and \ac{se} detectors \citep{entropyOCEANS}. \Ac{se} detectors have consistently shown better performance than energy detectors \citep{entropyJASA}. The previous approach by \citet{mypaper} modifies \ac{se} detectors to stabilize the entropy measure and provide pseudo-probabilistic output, thereby improving endpoint detection accuracy and the interpretability of the detector's output. \ac{se} detectors naturally motivate extension by including feature extraction and classification using the same \ac{tf}-decomposition used for \ac{se} calculation, ultimately reducing computational resources.

@@

Figure \ref{fig:det_roc} shows the \ac{roc} curve for detection on a sample-by-sample basis. Incremental performance improvement is demonstrated by adding a \ac{gmm} and enhancing the detector with whitening. In Figure \ref{fig:det_calls}, we show that the proposed detector is able to detect the majority calls while detecting fewer noise.  As illustrated in many previous studies, \ac{se} greatly outperforms a traditional \ac{bled} detector \citep{entropyJASA}. Whitening is also shown to greatly improve the capabilities of the \ac{bled} detector.

@@

As seen in Figures \ref{fig:det_roc} and \ref{fig:det_calls}, the proposed \ac{se} detector outperforms the previous method \citep{mypaper}. Typical \ac{bled} with a fixed threshold performs very poorly due to varying noise conditions. \ac{bled} has repeatedly been shown to not be a reliable \ac{poi} detector \citep{internalsignaldetection, entropyJASA}.

@@

\begin{table}[h!]
\caption{Number of labeled detections for all vocalisations of interest. The detections are produced by the whitened \ac{gmm} \ac{se} detector. A significant number of detections are discarded in the training data due to overlapping.}
\label{tab:numdetections}
% \begin{ruledtabular}
% \renewcommand{\arraystretch}{0.8}
\begin{tabular}{lc} 
\textbf{Vocalisation}      & \textbf{Number of detections}  \\ \hline
Bm\_Ant\_A & 1514              \\
Bm\_Ant\_B & 474              \\
Bm\_Ant\_Z & 91               \\
Bm\_D      & 328               \\
Bp & 50                \\
Noise      & 25638          \\
Overlapping & 150 \footnotemark[1]      \\ 
\end{tabular}
% \end{ruledtabular}
\footnotetext[1]{Overlapping detections contain at least 2 overlapping annotations of different classes. This implies that more than 300 vocalisations are discarded as a result of overlap, thereby drastically reducing training data for the \textit{Bm\_D} and \textit{Bp\_20Plus} classes, which have a higher tendency for overlap.}
\end{table}

@@

Our disposal of the time dimension by using the maximum naturally lends our classification model being replaced with a \ac{hmm} to incorporate time-dependent relationships similar to that of \citet{hmm2} and \citet{hmm1}. However, instead of classifying on a frame-by-frame basis such as in \citet{otherScattering, ws_audio, ws_ecg}, we utilise the scattering coefficients to first detect a variable-length segment of audio which may contain whale vocalisations, thereby reducing the amount of data in the noise class. This provides a distinct advantage over other methods, in that it already incorporates the extracted feature coefficients in detection prior to classification. As such, large portions of noise are discarded, the amount of which is tuned by the probability threshold of the proposed \ac{se} detector. This contrasts to \citet{casey2019}, which contains over 3 million noise class samples for 5000 vocalisation samples using the sliding window approach. This corresponds to a ratio of 1:600 for vocalisation to noise classes, whereas our method, which includes a detector, reduces this ratio to approximately 1:7 (all classes), at the cost of rejecting very low \ac{snr} vocalisations.

### 2024-05-06 dissertation/content/paper3/main.tex
consolodated references

Various forms of \ac{ws} exist, with \ac{1d} scattering first proposed by Anden and Mallat \citep{1dscattering1, 1dscattering2}, which was later extended to 2 \citep{2dscattering} and 3 dimensions \citep{3dscattering, harmonicscattering}. Additional and more specialised forms for \ac{ws} include rotation-invariant scattering \citep{groupinvariantscattering} for the classification of textures and joint-\ac{tf} scattering \citep{jointtfscattering1, jointtfscattering2}. Joint-\ac{tf} scattering utilises a separable \ac{2d} filter that operates on the first level scattering scalogram, denoted by the operator $\mathcal{U}_1$, which is the only apparent usage of separable filters in the current scattering literature.

In a deep-learning setting, sensible initialisation of filters prior to learning can significantly improve performance and interpretability \citep{sincnet}. The greatest advantage of separable filters is their computation speed when computing convolutions directly, as is performed in \acp{cnn} \citep{separablecnn}. Additionally, separable filters reduce the number of parameters of a \ac{cnn} if the filters are configured to be learnable. Learnable filters can typically improve performance compared to their fixed-filters counterpart \citep{learnablewavelettransform}.

>>>>

Various forms of \ac{ws} exist, with \ac{1d} scattering first proposed by Anden and Mallat \citep{1dscattering1, ws}, which was later extended to 2 \citep{2dscattering} and 3 dimensions \citep{3dscattering, harmonicscattering}. Additional and more specialised forms for \ac{ws} include rotation-invariant scattering \citep{groupinvariantscattering} for the classification of textures and joint-\ac{tf} scattering \citep{ws_joint_tf, jointtfscattering2}. Joint-\ac{tf} scattering utilises a separable \ac{2d} filter that operates on the first level scattering scalogram, denoted by the operator $\mathcal{U}_1$, which is the only apparent usage of separable filters in the current scattering literature.

In a deep-learning setting, sensible initialisation of filters prior to learning can significantly improve performance and interpretability \citep{sincnet}. The greatest advantage of separable filters is their computation speed when computing convolutions directly, as is performed in \acp{cnn} \citep{separablecnn}. Additionally, separable filters reduce the number of parameters of a \ac{cnn} if the filters are configured to be learnable. Learnable filters can typically improve performance compared to their fixed-filters counterpart \citep{scattering_birdsong}.

### 2024-05-06 dissertation/content/paper3/original/main.tex
consolodated references

Various forms of \ac{ws} exist, with \ac{1d} scattering first proposed by Anden and Mallat \cite{1dscattering1, 1dscattering2}, which was later extended to 2 \cite{2dscattering} and 3 dimensions \cite{3dscattering, harmonicscattering}. Additional and more specialised forms for \ac{ws} include rotation-invariant scattering \cite{groupinvariantscattering} for the classification of textures and joint-\ac{tf} scattering \cite{jointtfscattering1, jointtfscattering2}. Joint-\ac{tf} scattering utilises a separable \ac{2d} filter that operates on the first level scattering scalogram, denoted by the operator $\mathcal{U}_1$, which is the only apparent usage of separable filters in the current scattering literature.

In a deep-learning setting, sensible initialisation of filters prior to learning can significantly improve performance and interpretability \cite{sincnet}. The greatest advantage of separable filters is their computation speed when computing convolutions directly, as is performed in \acp{cnn} \cite{separablecnn}. Additionally, separable filters reduce the number of parameters of a \ac{cnn} if the filters are configured to be learnable. Learnable filters can typically improve performance compared to their fixed-filters counterpart \cite{learnablewavelettransform}.

>>>>

Various forms of \ac{ws} exist, with \ac{1d} scattering first proposed by Anden and Mallat \cite{1dscattering1, ws}, which was later extended to 2 \cite{2dscattering} and 3 dimensions \cite{3dscattering, harmonicscattering}. Additional and more specialised forms for \ac{ws} include rotation-invariant scattering \cite{groupinvariantscattering} for the classification of textures and joint-\ac{tf} scattering \cite{ws_joint_tf, jointtfscattering2}. Joint-\ac{tf} scattering utilises a separable \ac{2d} filter that operates on the first level scattering scalogram, denoted by the operator $\mathcal{U}_1$, which is the only apparent usage of separable filters in the current scattering literature.

In a deep-learning setting, sensible initialisation of filters prior to learning can significantly improve performance and interpretability \cite{sincnet}. The greatest advantage of separable filters is their computation speed when computing convolutions directly, as is performed in \acp{cnn} \cite{separablecnn}. Additionally, separable filters reduce the number of parameters of a \ac{cnn} if the filters are configured to be learnable. Learnable filters can typically improve performance compared to their fixed-filters counterpart \cite{scattering_birdsong}.
